---
params:
  handout: 
title: "SM-2302 Software for Mathematicians"
subtitle: "R1: Logic and Types in R \\only<handout>{\\emph{[handout version]}}"
author: "Dr. Haziq Jamil"
date: "Semester I 2023/24"
institute: |
  | Mathematical Sciences, Faculty of Science, UBD
  | \url{https://haziqj.ml}
output: 
  beamer_presentation:
    template: ubd_beamer_rmd.tex
    latex_engine: xelatex
    slide_level: 3
    keep_tex: false
    citation_package: biblatex
    pandoc_args: ["--lua-filter=luafilters/mycustom.lua", "--lua-filter=luafilters/bookdown.lua"] 
header-includes:
  - \renewcommand{\Pr}{\operatorname{P}}
  - \usetikzlibrary{fit}
  - \usetikzlibrary{shapes.geometric}
# include-after:
#   -
aspectratio: 169
handout: "`r params$handout`"
toc: false
toctitle: Overview
tocmulticol:
transitions: true
progressdots: false
# banner: true
# logo: true
# bibliography: 
# bibstyle: 
refslide: false
thankyou: false
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})
# https://stackoverflow.com/questions/25646333/code-chunk-font-size-in-rmarkdown-with-knitr-and-latex

knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE,
  message = FALSE,
  error = TRUE,
  fig.height = 5, 
  fig.width = 5, 
  fig.path = "figure/", 
  fig.align = "center",
  cache = TRUE, 
  cache.path = "_cache/",
  size = "footnotesize"
)
options(width = 85)  # if 4:3 set to 55; if 16:9 set to 70
library(tidyverse)
theme_set(theme_classic())
library(tidyverse)
library(ggpubr)
library(cowplot)
library(ggrepel)
library(ggforce)

ubdblue <- "#325494"
ubdteal <- "#58DDB3"
ubdyellow <- "#F5C946"
ubdred <- "#B10F2E"
navyblue <- "#002f5c"
solidpink <- "#8E3B46"

mycol <- c("#30123BFF", "#4777EFFF", "#1BD0D5FF", "#62FC6BFF", "#D2E935FF",
           "#FE9B2DFF", "#DB3A07FF", "#7A0403FF")
```



### Introduction

> In R (almost) everything is a vector


\vspace{1em}

The fundamental building block of data in R are vectors (collections of related values, objects, data structures, etc). 

\vspace{1em}

R has two types of vectors:

* **atomic** vectors (*vectors*)

    - homogeneous collections of the *same* type (e.g. all `TRUE`/`FALSE` values, all numbers, or all character strings).

* **generic** vectors (*lists*)
  
    - heterogeneous collections of *any* type of R object, even other lists <br/> 
    (meaning they can have a hierarchical/tree-like structure).

\blfootnote{R material lecture slides largely based off \url{https://sta323-sp22.github.io/}}


# Atomic Vectors

R has six atomic vector types, we can check the type of any object in R using the `typeof()` function

`typeof()`  |  `mode()`     
:-----------|:------------
logical     |  logical    
double      |  numeric    
integer     |  numeric    
character   |  character  
complex     |  complex    
raw         |  raw        

Mode is a higher level abstraction, we will discuss this in detail a bit later.

<!-- .footnote[ -->
<!-- There are additional types in R, e.g. generic vectors have type `list`, but more on these later. See `?typeof` for more information. -->
<!-- ] -->


### `logical` -- Boolean values (`TRUE` and `FALSE`)

\vspace{-1em}

::: {.columns}

:::: {.column width=48%}
```{r}
typeof(TRUE)
typeof(FALSE)
```
:::

:::: {.column width=48%}
```{r}
mode(TRUE)
mode(FALSE)
```
:::

:::

\vspace{1em}

R will let you use `T` and `F` as shortcuts to `TRUE` and `FALSE`, this is a bad practice as these values are actually global variables that can be overwritten.

```{r}
T
T <- FALSE
T
```


### `character` -- text strings

Either single or double quotes are fine, opening and closing quote must match.

\vspace{-1em}

::: {.columns}

:::: {.column width=48%}
```{r}
typeof("hello")
typeof('world')
```
::::

:::: {.column width=48%}
```{r}
mode("hello")
mode('world')
```
::::

:::

\vspace{1em}

Quote characters can be included by escaping or using a non-matching quote.

```{r}
"abc'123"
'abc"123'
```


### Numeric types

`double` - floating point values (these are the default numerical type)

\vspace{-1em}

::: {.columns}

:::: {.column width=48%}
```{r}
typeof(1.33)
typeof(7)
```
::::

:::: {.column width=48%}
```{r}
mode(1.33)
mode(7)
```
::::

:::

\vspace{1em}

`integer` - integer values (literals are indicated with an `L` suffix)

\vspace{-1em}

::: {.columns}

:::: {.column width=48%}
```{r}
typeof( 7L )
typeof( 1:3 )
```
::::

:::: {.column width=48%}
```{r}
mode( 7L )
mode( 1:3 )
```
::::

:::

### Concatenation

Atomic vectors can be grown (combined) using the concatenate `c()` function.

```{r}
c(1, 2, 3)
c("Hello", "World!")
c(1, 1:10)
c(1, c(2, c(3)))
```

\blfootnote{Note: Atomic vectors are inherently flat.}


### Inspecting types

- `typeof(x)`: returns a character vector (length 1) of the *type* of object `x`.

- `mode(x)`: returns a character vector (length 1) of the *mode* of object `x`.

- `str(x)`: compactly display the internal *str*ucture of object `x`.

\vspace{-1em}

::: {.columns}

:::: {.column width=32%}
```{r}
typeof(1)
typeof(1L)
typeof("A")
typeof(TRUE)
```
::::

:::: {.column width=32%}
```{r}
mode(1)
mode(1L)
mode("A")
mode(TRUE)
```
::::

:::: {.column width=32%}
```{r}
str(1)
str(1L)
str("A")
str(TRUE)
```
::::

:::


### Type predicates

* `is.logical(x)`   - returns `TRUE` if `x` has *type* `logical`.
* `is.character(x)` - returns `TRUE` if `x` has *type* `character`.
* `is.integer(x)`   - returns `TRUE` if `x` has *type* `integer`.
* `is.numeric(x)`   - returns `TRUE` if `x` has *mode* `numeric`.

\vspace{-1em}

::: {.columns}

:::: {.column width=32%}
```{r}
is.integer(1)
is.integer(1L)
is.integer(3:7)
```
::::

:::: {.column width=32%}
```{r}
is.double(1)
is.double(1L)
is.double(3:8)
```
::::

:::: {.column width=32%}
```{r}
is.numeric(1)
is.numeric(1L)
is.numeric(3:7)
```
::::

:::

\vspace{1em}


Many other useful predicates: `is.double()`, `is.atomic()`, `is.list()`, `is.vector()`, and some packages provide their own too.


### Type coercion

R is a dynamically typed language -- it will automatically convert between most types without raising warnings or errors. Keep in mind the rule that atomic vectors must always contain values of the same type.

```{r}
c(1, "Hello")
c(FALSE, 3L)
c(1.2, 3L)
```

### Operator coercion

Operators and functions will generally attempt to coerce values to an appropriate type for the given operation.

\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
3.1+1L
5 + FALSE
TRUE & FALSE
TRUE & 7
```
::::

:::: {.column width=49%}
```{r}
log(1)
log(TRUE)
TRUE | FALSE
FALSE | !5
```
::::

:::


### Explicit coercion

Most of the `is` functions we just saw have an `as` variant which can be used for *explicit* coercion.

\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
as.logical(5.2)
as.character(TRUE)
as.integer(pi)
```
::::

:::: {.column width=49%}
```{r}
as.numeric(FALSE)
as.double("7.2")
as.double("one")
```
::::

:::

# Conditionals & Control Flow

### Logical (boolean) operators



|  Operator                     |  Operation    |  Vectorized? 
|:-----------------------------:|:-------------:|:------------:
| <code>x &#124; y</code>       |  or           |   Yes        
| `x & y`                       |  and          |   Yes        
| `!x`                          |  not          |   Yes        
| <code>x &#124;&#124; y</code> |  or           |   No         
| `x && y`                      |  and          |   No         
|`xor(x, y)`                    |  exclusive or |   Yes     


### Vectorized?

```{r}
x <- c(TRUE, FALSE, TRUE)
y <- c(FALSE, TRUE, TRUE)
```

\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
x | y
x & y
```
::::

:::: {.column width=49%}
```{r}
x || y
x && y
```
::::

:::

\vspace{2em}

Note: both `||` and `&&` only use the *first* value in the vector, all other values are ignored, there is no warning about the ignored values.


### Vectorization and math

Almost all of the basic mathematical operations (and many other functions) in R are vectorized.

\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
c(1, 2, 3) + c(3, 2, 1)
c(1, 2, 3) / c(3, 2, 1)
```
::::

:::: {.column width=49%}
```{r}
log(c(1, 3, 0))
sin(c(1, 2, 3))
```
::::

:::



### Length coercion (aka recycling)

```{r}
x <- c(TRUE, FALSE, TRUE)
y <- TRUE
z <- c(FALSE, TRUE)
```

\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
x | y
x & y
```
::::

:::: {.column width=49%}
```{r}
y | z
y & z
```
::::

:::

\vspace{1em}

```{r}
x | z
```


### Length coercion and math

The same length coercion rules apply for most basic mathematical operators as well.



```{r}
x <- c(1, 2, 3)
y <- c(5, 4)
z <- 10L
```

\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
x + x
x + z
```
::::

:::: {.column width=49%}
```{r}
log(x)
y / z
```
::::

:::


```{r}
x %% y
```



### Comparison operators

  Operator  |  Comparison                |  Vectorized?
:----------:|:--------------------------:|:----------------:
 `x < y`    |  less than                 |  Yes
 `x > y`    |  greater than              |  Yes
 `x <= y`   |  less than or equal to     |  Yes
 `x >= y`   |  greater than or equal to  |  Yes
 `x != y`   |  not equal to              |  Yes
 `x == y`   |  equal to                  |  Yes
 `x %in% y` |  contains                  |  Yes (over `x`)\footnotemark
 
 
 \footnotetext{Over `x` here means the returned value will have the same length as `x`.}
 
### Comparisons

```{r}
x <- c("A","B","C")
z <- "A"
```

\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
x == z
x != z
x > z
```
::::

:::: {.column width=49%}
```{r}
x %in% z
z %in% x
```
::::

:::

 
### Conditional control flow

Conditional execution of code blocks is achieved via `if` statements. 

```{r}
x <- c(1, 3)
```


\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
if (3 %in% x)
  print("Contains 3!")
if (1 %in% x)
  print("Contains 1!")
```
::::

:::: {.column width=49%}
```{r}
if (5 %in% x)
  print("Contains 5!")
```

```{r}
if (5 %in% x) {
  print("Contains 5!")
} else {
  print("Does not contain 5!")
}
```
::::

:::

### `if` is not vectorized

```{r}
x <- c(1, 3)
```



```{r, error = TRUE}
if (x == 1)
  print("x is 1!")
```


```{r, error = TRUE}
if (x == 3)
  print("x is 3!")
```


### Collapsing logical vectors

There are a couple of helper functions for collapsing a logical vector down to a single value: `any`, `all`

```{r}
x <- c(3,4,1)
```

::: {.columns}

:::: {.column width=49%}
```{r}
x >= 2
any(x >= 2)
all(x >= 2)
x <= 4
```
::::

:::: {.column width=49%}


```{r}
any(x <= 4)
all(x <= 4)
if (any(x == 3)) 
  print("x contains 3!")
```
::::

:::



### `else if` and `else`

::: {.columns}

:::: {.column width=49%}
```{r}
x <- 3

if (x < 0) {
  "x is negative"
} else if (x > 0) {
  "x is positive"
} else {
  "x is zero"
}
```
::::

:::: {.column width=49%}


```{r}
x <- 0

if (x < 0) {
  "x is negative"
} else if (x > 0) {
  "x is positive"
} else {
  "x is zero"
}
```
::::

:::



### `if` and `return`

R's `if` conditional statements return a value (invisibly), the two following implementations are equivalent.

::: {.columns}

:::: {.column width=49%}
```{r}
x <- 5
```

```{r}
s <- if (x %% 2 == 0) {
  x / 2
} else {
  3 * x + 1
}
```

```{r}
s
```
::::

:::: {.column width=49%}


```{r}
x <- 5
```

```{r}
if (x %% 2 == 0) {
  s <- x / 2
} else {
  s <- 3 * x + 1
}
```

```{r}
s
```
::::

:::

\vspace{1em}

Notice that conditional expressions are evaluated in the parent scope.

# Error Checking


### `stop` and `stopifnot`

Often we want to validate user input or function arguments - if our assumptions are not met then we often want to report the error and stop execution. 



```{r error=TRUE}
ok <- FALSE
```

```{r error=TRUE}
if (!ok)
  stop("Things are not ok.")
```

```{r error=TRUE}
stopifnot(ok)
```




### Style choices

::: {.columns}

:::: {.column width=49%}

Do stuff:
```{r eval=FALSE}
if (condition_one) {
  ##
  ## Do stuff
  ##
} else if (condition_two) {
  ##
  ## Do other stuff
  ##
} else if (condition_error) {
  stop("Condition error occured")
}
```
::::


:::: {.column width=49%}
Do stuff (better):
```{r eval=FALSE}
# Do stuff better
if (condition_error) {
  stop("Condition error occured")
}

if (condition_one) {
  ##
  ## Do stuff
  ##
} else if (condition_two) {
  ##
  ## Do other stuff
  ##
}
```

::::

:::

# Missing Values


R uses `NA` to represent missing values in its data structures, what may not be obvious is that there are different `NA`s for different atomic types.

\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
typeof(NA)
typeof(NA + 1)
typeof(NA + 1L)
typeof(c(NA, ""))
```
::::


:::: {.column width=49%}
```{r}
typeof(NA_character_)
typeof(NA_real_)
typeof(NA_integer_)
typeof(NA_complex_)
```

::::

:::

### NA "stickiness" 

Because `NA`s represent missing values it makes sense that any calculation using them should also be missing.

\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
1 + NA
1 / NA
NA * 5
```
::::


:::: {.column width=49%}
```{r}
sqrt(NA)
3 ^ NA
sum(c(1, 2, 3, NA))
```

::::

:::

\vspace{0.5em}

Summarizing functions (e.g. `sum()`, `mean()`, `sd()`, etc.) will often have a `na.rm` argument which will allow you to *drop* missing values.

\vspace{-0.5em}

```{r}
sum(c(1, 2, 3, NA), na.rm = TRUE)
```



### NAs are not always sticky 
  
A useful mental model for `NA`s is to consider them as a unknown value that could take any of the possible values for that type. 
For numbers or characters this isn't very helpful, but for a logical value we know that the value must either be `TRUE` or `FALSE` and we can use that when deciding what value to return.


```{r}
TRUE & NA
```

```{r}
FALSE & NA
```

```{r}
TRUE | NA
```

```{r}
FALSE | NA
```




### Conditionals and missing values

`NA`s can be problematic in some cases (particularly for control flow)



```{r error=TRUE}
1 == NA
```


```{r error=TRUE}
if (2 != NA)
  "Here"
```


```{r error=TRUE}
if (all(c(1, 2, NA, 4) >= 1))
  "There"
```


```{r error=TRUE}
if (any(c(1, 2, NA, 4) >= 1))
  "There"
```


### Testing for `NA`

To explicitly test if a value is missing it is necessary to use `is.na` (often along with `any` or `all`).


::: {.columns}

:::: {.column width=49%}
```{r}
NA == NA
is.na(NA)
is.na(1)
```
::::


:::: {.column width=49%}
```{r}
is.na(c(1, 2, 3, NA))
any(is.na(c(1, 2, 3, NA)))
all(is.na(c(1, 2, 3, NA)))
```

::::

:::

### Other special values (double)

These are defined as part of the IEEE floating point standard (not unique to R)

* `NaN` -- Not a number
* `Inf` -- Positive infinity
* `-Inf` -- Negative infinity


::: {.columns}

:::: {.column width=49%}
```{r}
pi / 0
0 / 0
1 / 0 + 1 / 0
```
::::


:::: {.column width=49%}
```{r}
1 / 0 - 1 / 0
NaN / NA
NaN * NA
```

::::

:::



### Testing for `Inf` and `NaN`

`NaN` and `Inf` don't have the same testing issues that `NA`s do, but there are still convenience functions for testing for these types of values

\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
is.finite(Inf)
is.infinite(-Inf)
is.nan(Inf)
is.nan(-Inf)
Inf > 1
-Inf > 1
```
::::


:::: {.column width=49%}
```{r}
is.finite(NaN)
is.infinite(NaN)
is.nan(NaN)
is.finite(NA)
is.infinite(NA)
is.nan(NA)
```

::::

:::


### Coercion for infinity and NaN
  
First remember that `Inf`, `-Inf`, and `NaN` are doubles, however their coercion behavior is not the same as for other doubles

```{r}
as.integer(Inf)
as.integer(NaN)
```

\vspace{-1em}

::: {.columns}

:::: {.column width=49%}
```{r}
as.logical(Inf)
as.logical(NaN)
```
::::


:::: {.column width=49%}
```{r}
as.character(Inf)
as.character(NaN)
```

::::

:::



# Functions


### Function parts

Functions are defined by two components: the arguments (`formals`) and the code (`body`). 
Functions are assigned names like any other object in R (using `=` or `<-`)

```{r}
gcd <- function(x1, y1, x2 = 0, y2 = 0) {
  R <- 6371  # Earth mean radius in km
  acos(sin(y1) * sin(y2) + cos(y1) * cos(y2) * cos(x2 - x1)) * R   # distance in km
}
```

\vspace{-2em}

::: {.columns}

:::: {.column width=25%}
```{r}
typeof(gcd)
formals(gcd)
```
::::


:::: {.column width=72%}
```{r}
mode(gcd)
body(gcd)
```

::::

:::


### Return values

There are two approaches to returning values from functions in R.

\vspace{1em}

**Explicit**: using one or more `return` function calls

```{r}
f <- function(x) {
  return(x * x)
}
f(2)
```



**Implicit**: return value of the last expression is returned.

```{r}
g <- function(x) {
  x * x
}
g(3)
```



<!-- ### Invisible returns -->

<!-- Many functions in R make use of an invisible return value -->

<!-- ```{r} -->
<!-- f = function(x) { -->
<!--   print(x) -->
<!-- } -->

<!-- y = f(1) -->
<!-- y -->
<!-- ``` -->



<!-- ```{r} -->
<!-- g = function(x) { -->
<!--   invisible(x) -->
<!-- } -->
<!-- ``` -->

<!-- ```{r} -->
<!-- g(2) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- z = g(2) -->
<!-- z -->
<!-- ``` -->



### Returning multiple values

If we want a function to return more than one value we can group things using atomic vectors or lists.

```{r}
f <- function(x) c(x, x ^ 2, x ^ 3)
f(1:2)
```

```{r}
g <- function(x) list(x, "hello")
g(1:2)
```

More on lists next time.

### Argument names

When defining a function we explicitly define names for the arguments, which become variables within the scope of the function.
When calling a function we can use these names to pass arguments in an alternative order.


```{r}
f <- function(x, y, z) {
  paste0("x = ", x, ", y = ", y, ", z = ", z)
}
```

\vspace{-3em}

::: {.columns}

:::: {.column width=48%}
```{r}
f(1, 2, 3)
f(z = 1, x = 2, y = 3)
```
::::


:::: {.column width=48%}
```{r}
f(y = 2, 1, 3)
f(y = 2, 1, x = 3)
```

::::

:::

```{r, error=TRUE}
f(1, 2, 3, 4)
```



### Argument defaults

It is also possible to give function arguments default values, so that they don't need to be provided every time the function is called.

```{r}
f <- function(x, y = 1, z = 1) {
  paste0("x = ", x, ", y = ", y, ", z = ", z)
}
```

\vspace{-3em}

::: {.columns}

:::: {.column width=48%}
```{r}
f(3)
f(x = 3)
```
::::


:::: {.column width=48%}
```{r}
f(z = 3, x = 2)
f(y = 2, 2)
```

::::

:::

```{r, error=TRUE}
f()
```



### Scope

R has generous scoping rules, if it can't find a variable in the current scope (e.g. a function's body) it will look for it in the next higher scope, and so on.

::: {.columns}

:::: {.column width=48%}
```{r}
y <- 1

f <- function(x) {
  x + y
}

f(3)
```
::::


:::: {.column width=48%}
```{r}
y <- 1

g <- function(x) {
  y <- 2
  x + y
}

g(3)
y
```

::::

:::

### Scope (cont.)


Additionally, variables defined within a scope only persist for the duration of that scope, and do not overwrite variables at a higher scope.

```{r}
x <- y <- z <- 1

f <- function() {
    y <- 2
    g <- function() {
      z <- 3
      return(x + y + z)
    }
    return(g())
}
f()
c(x, y, z)
```




# Loops


### `for` loops

Simplest, and most common type of loop in R--given a vector iterate through the elements and evaluate the code block for each.


```{r}
is_even <- function(x) {
  res <- c()
  
  for(val in x) {
    res <- c(res, val %% 2 == 0)
  }
  
  res
}
is_even(1:10)
is_even(seq(1, 5, 2))
```


### `while` loops

Repeat until the given condition is **not** met (i.e. evaluates to `FALSE`)

```{r}
make_seq <- function(from = 1, to = 1, by = 1) {
  res <- c(from)
  cur <- from
  
  while(cur + by <= to) {
    cur = cur + by
    res = c(res, cur)
  }
  
  res
}
make_seq(1, 6)
make_seq(1, 6, 2)
```

<!-- ### `repeat` loops -->

<!-- Repeat the loop until a `break` is encountered -->

<!-- ```{r} -->
<!-- make_seq2 = function(from = 1, to = 1, by = 1) { -->
<!--   res <- c(from) -->
<!--   cur <- from -->

<!--   repeat { -->
<!--     cur <- cur + by -->

<!--     if (cur > to) -->
<!--       break -->

<!--     res <- c(res, cur) -->
<!--   } -->

<!--   res -->
<!-- } -->
<!-- make_seq2(1, 6) -->
<!-- make_seq2(1, 6, 2) -->
<!-- ``` -->



### Some helpful functions

Often we want to use a loop across the indexes of an object and not the elements themselves. There are several useful functions to help you do this: `:`, `length`, `seq`, `seq_along`, `seq_len`, etc.


::: {.columns}

:::: {.column width=48%}
```{r}
4:7
length(4:7)
seq(4,7)
```
::::


:::: {.column width=48%}
```{r}
seq_along(4:7)
seq_len(length(4:7))
seq(4, 7, by = 2)
```

::::

:::




### Avoid using `1:length(x)`

A common loop construction you'll see in a lot of R code is using `1:length(x)` to generate a vector of index values for the vector `x`. 

\vspace{-1em}

::: {.columns}

:::: {.column width=48%}
```{r}
f <- function(x) {
  for(i in 1:length(x)) {
    print(i)
  }
}
f(2:1)
f(2)
f(integer())
```
::::


:::: {.column width=48%}
```{r}
g <- function(x) {
  for(i in seq_along(x)) {
    print(i)
  }
}
g(2:1)
g(2)
g(integer())
```

::::

:::

